---
title: "Exploration of Air Quality Dataset"
author: "Shailesh Kumar"
output: 
  html_notebook:
    toc: true
    number_sections: true
---
This data set has daily air quality measurements in New York from
May to September 1973 over a period of 5 months. 

| Variable | Type | Unit | Description | Detail |
|----------|------|------|-------------|--------|
| Ozone | numeric | parts per billion | mean Ozone concentration |  Roosevelt island |
| Solar.R | numeric |  | Solar radiaion | central park | 
| Wind | numeric | miles per hour | average wind speed | LaGuardia airport |
| Temp |numeric | Fahrenheit | maximum daily temperature | LaGuardia airport |
| Month | numeric |  | Month of observation | 5 values |
| Day |  numeric |  | day of month | 153 days total |
|  | | | | |


Necessary packages
```{r}
require(ggplot2)
require(moments)
```
Let's make a copy of the dataset for the 
following analysis. 
```{r}
aq <- data.frame(airquality)
```


Names of columns
```{r}
colnames(aq)
```


Let's look at some of the rows:

```{r}
head(aq)
```

Several entries are not available. Let's summarize:

```{r}
colSums(is.na(aq))
```


Day should be a factor variable
```{r}
aq$Day <- factor(aq$Day, levels=c(1:31), ordered=TRUE)
```
Month should be a factor variable
```{r}
aq$Month <- factor(aq$Month, levels=5:9, labels=month.abb[5:9], ordered=TRUE)
```


Summary statistiscs for all the variables:
```{r}
summary(aq)
```

# Variables

## Ozone

```{r}
summary(aq$Ozone)
```

Histogram
```{r}
ggplot(data=aq) + geom_histogram(mapping=aes(Ozone))
```

The histogram appears assymetric. 


We can check the normalcy of the distribution by measuring its 
skewness and kurtosis

```{r}
skewness(aq$Ozone, na.rm = T)
```
The data is skewed heavily towards right.

Let's also measure kurtosis:
```{r}
kurtosis(aq$Ozone, na.rm=T)
```
This is *leptokurtic*.

Let's apply a transformation to see if the histogram becomes more symmetric:


```{r}
aq$Ozone.sqrt <- sqrt(aq$Ozone)
ggplot(data=aq) + geom_histogram(mapping=aes(Ozone.sqrt))
```
We have stored the square root of Ozone concentration values in 
the aq dataframe itself as a new column.
This is indeed more symmetric and looking like normal distribution.

Let's verify the skewness:
```{r}
skewness(aq$Ozone.sqrt, na.rm=T)
```
Also the kurtosis:
```{r}
kurtosis(aq$Ozone.sqrt, na.rm=T)
```

There is indeed improvement in skewness.

Seasonal variation in data:
```{r}
qplot(Month, Ozone, data=aq, geom="boxplot", color=Month)
```


## Solar.R

```{r}
summary(aq$Solar.R)
```

Histogram
```{r}
ggplot(data=aq) + geom_histogram(mapping=aes(Solar.R))
```

Seasonal variation in data:
```{r}
qplot(Month, Solar.R, data=aq, geom="boxplot", color=Month)
```

## Wind

```{r}
summary(aq$Wind)
```


Histogram
```{r}
ggplot(data=aq) + geom_histogram(mapping=aes(Wind))
```

Let's verify the normalcy of the distribution
```{r}
skewness(aq$Wind)
kurtosis(aq$Wind)
```
This is very close to a normal distribution.

Seasonal variation in data:
```{r}
qplot(Month, Wind, data=aq, geom="boxplot", color=Month)
```

## Temp

```{r}
summary(aq$Temp)
```

Seasonal variation in data:
```{r}
qplot(Month, Temp, data=aq, geom="boxplot", color=Month)
```

## Month

```{r}
table(aq$Month)
```

This just reflects the number of days in each month from May to September. 

## Day

```{r}
table(aq$Day)
```

As expected, day 31 appears only 3 times. 


# Correlations

Let us find out the Pearson correlation coefficients between
the different variables in the dataset:

```{r}
# Only numerical variables can be used for correlation
columns <- c('Ozone.sqrt', 'Solar.R', 'Wind', 'Temp')
# We identify non-na rows as rows with na values cause problems 
rows <- rowSums(is.na(aq)) == 0
# Compute correlation of variables and round to 2 digits
round(cor(aq[rows, columns]), 2)
```

* Solar Radiation is moderately correlated with ozone concentration.
* wind and ozone seem strongly correlated.
* Temperature and ozone are strongly correlated. 
* Very weak correlation between wind and solar radiation.
* Temperature and solar radition are weakly correlated.
* Temperature and wind are moderately correlated.


Let us verify these correlations using scatter plots.


##Very weak correlation: 
```{r}
qplot(Solar.R, Wind, data=aq, geom="point", color=Month)
```

##Weak correlation:
```{r}
qplot(Temp, Solar.R, data=aq, geom="point", color=Month)
```

##Moderate correlation:
```{r}
qplot(Solar.R, Ozone.sqrt, data=aq, color=Month, geom="point")
```
```{r}
qplot(Temp, Wind, data=aq, color=Month, geom="point")
```

##Strong correlation
```{r}
qplot(Wind, Ozone.sqrt, data=aq, color=Month, geom='point')
```

```{r}
qplot(Temp, Ozone.sqrt, data=aq, color=Month, geom='point')
```

