---
title: "Exploration of Diamonds Dataset"
author: "Shailesh Kumar"
output: 
  html_notebook:
    toc: true
    number_sections: true
---

This notebook explores the diamonds dataset shipped with the ggplot2 package.

The diamonds dataset contains the price and other attributes like
carat, cut, clarity, etc. of 53940 diamonds. 

| Variable | Type | Description | Details |
|----------|------|-------------|---------|
| carat | numeric | weight of the diamond | 0.2 - 5.01  |
| cut | ordinal | quality of the cut |  Fair      Good Very Good   Premium     Ideal| 
| color | ordinal | diamond color | D     E     F     G     H     I     J | 
| clarity | ordinal| clarity of the diamond | I1   SI2   SI1   VS2   VS1  VVS2  VVS1    IF | 
| depth | numeric | total depth percentage | 43-79 | 
| table | numeric | width of top of diamond relative to widest point | 43-95 | 
| price | numeric | price in US dollars |
| x | numeric | length in mm | 0-10.74 | 
| y | numeric | width in mm | 0-58.9 | 
| z | numeric | depth in mm | 0-31.8 | 


Let's load essential libraries:
```{r}
require(ggplot2)
# provides gam method for geom smooth
require(mgcv)
# provides natural splines for the lm method
require(splines)
```

Number of variables:
```{r}
ncol(diamonds)
```



Size of the dataset:
```{r}
nrow(diamonds)
```


Let's see some rows from the dataset:
```{r}
head(diamonds)
```

# Variables

## carat

```{r}
summary(diamonds$carat)
```

```{r}
qplot(x=carat, data=diamonds, binwidth=1)
qplot(x=carat, data=diamonds, binwidth=0.1)
qplot(x=carat, data=diamonds, binwidth=0.01)
```

```{r}
qplot(x=carat, data=diamonds, binwidth=0.1, fill=color)
qplot(x=carat, data=diamonds, geom='density', color=color)
```


## cut

```{r}
table(diamonds$cut)
```

```{r}
qplot(cut, data=diamonds, geom='bar', fill=cut)
```

## color

```{r}
table(diamonds$color)
```

```{r}
qplot(color, data=diamonds, geom='bar', fill=color)
```

## clarity

```{r}
table(diamonds$clarity)
```

```{r}
qplot(clarity, data=diamonds, geom='bar', fill=clarity)
```

## depth

```{r}
summary(diamonds$depth)
```

```{r}
qplot(depth, data=diamonds)
```

## table

```{r}
summary(diamonds$table)
```

```{r}
qplot(table, data=diamonds, geom='histogram')
```

## price

```{r}
summary(diamonds$price)
```

```{r}
qplot(price, data=diamonds, geom='histogram')
```

Price by cut:
```{r}
qplot(cut, price, data=diamonds, geom='boxplot', color=cut)
```

```{r}
qplot(cut, price / carat, data=diamonds, geom='boxplot', color=cut)
```


Price by clarity:
```{r}
qplot(x=clarity, y=price, color=clarity, data=diamonds, geom='boxplot')
```

```{r}
qplot(x=clarity, y=price / carat, color=clarity, data=diamonds, geom='boxplot')
```

Price by color
```{r}
qplot(color, price, data=diamonds, color=color, geom='boxplot')
```

```{r}
qplot(color, price/carat, data=diamonds, color=color, geom='boxplot')
```
```{r}
qplot(color, price, data=diamonds, color=color, geom='jitter', alpha=I(1/20))
```


## x

```{r}
summary(diamonds$x)
```

```{r}
qplot(x, data=diamonds, geom='histogram')
```

## y

```{r}
summary(diamonds$y)
```

```{r}
qplot(y, data=diamonds, geom='histogram')
```

## z

```{r}
summary(diamonds$z)
```

```{r}
qplot(z, data=diamonds, geom='histogram')
```

```{r}
qplot(log(1+z), data=diamonds, geom='histogram')
```

# Correlations


## carat vs price

```{r}
qplot(carat, price, data=diamonds, color=clarity, alpha=I(1/10))
```

```{r}
qplot(log(carat), log(price), data=diamonds, color=clarity, alpha=I(1/5))
```

```{r}
qplot(data=diamonds, x=carat, y=price, geom=c('point', 'smooth'))
```


```{r}
qplot(data=diamonds, x=carat, y=price, geom=c('point', 'smooth'), 
      method='gam', formula=y~s(x))
```

```{r}
# fitting a linear model with a natural splines formula
qplot(data=diamonds, x=carat, y=price, geom=c('point', 'smooth'), 
      method='lm', formula=y~ns(x,5))
```

## weight vs volume

```{r}
qplot(carat, x * y * z, data=diamonds)
```


## cut vs color

```{r}
qplot(x=cut, data=diamonds, geom='bar', fill=color)
```

## cut vs clarity

```{r}
qplot(data=diamonds, x=cut, fill=clarity, geom='bar')
```

## clarity vs color

```{r}
qplot(data=diamonds, x=clarity, fill=color, geom='bar')
```

