---
title: "Exploration of IRIS Dataset"
author: "Shailesh Kumar"
output: 
  html_notebook:
    toc: true
    number_sections: true
---
# Overview

This dataset contains the measurements of the variables 
sepal width and length, petal width and length for 50 flowers
from 3 species of iris.

Necessary packages:

```{r}
require(ggplot2)
```

```{r}
str(iris)
```


```{r}
head(iris)
```

# Variables

## Sepal.Length

```{r}
summary(iris$Sepal.Length)
```

Histogram of sepal length
```{r}
qplot(Sepal.Length, data=iris, geom='histogram', fill=Species, alpha=I(1/2))
```


## Sepal.Width

```{r}
summary(iris$Sepal.Width)
```

## Petal.Length


```{r}
summary(iris$Petal.Length)
```

Histogram of petal length
```{r}
qplot(Petal.Length, data=iris, geom='histogram')
```

Estimating a kernel denstiy for the distribution of petal length:
```{r}
qplot(Petal.Length, data=iris, geom='density')
```

Separating the petal length histogram on the basis of species:

```{r}
qplot(Petal.Length, data=iris, geom='histogram', fill=Species, alpha=I(0.5))
```

Densities on the basis of species
```{r}
qplot(Petal.Length, data=iris, geom='density', color=Species, fill=Species)
```


## Petal.Width

```{r}
summary(iris$Petal.Width)
```



# Correlations

## With Species

```{r}
cor(iris$Sepal.Length, as.numeric(iris$Species), method='spearman')
```

```{r}
cor(iris$Sepal.Width, as.numeric(iris$Species), method='spearman')
```

```{r}
cor(iris$Petal.Length, as.numeric(iris$Species), method='spearman')
```
```{r}
cor(iris$Petal.Length, as.numeric(iris$Species), method='spearman')
```


# Clustering

## K-means 

We note that petal length and width are very strongly correlated with species.
Are these two numbers enough to correctly predict the species? 

Let us prepare a data frame with only these two columns:
```{r}
iris2 <- iris[, c("Petal.Length", "Petal.Width")]
```

Let us perform k-means clustering on this data set
```{r}
num_clusters <- 3
set.seed(1111)
result <- kmeans(iris2, num_clusters, nstart=20)
```

The k-means algorithm has identified 3 centers:
```{r}
result$centers
```

The cluster assignments are as follows:
```{r}
result$cluster
```

Mapping between cluster assignment and original species labels:
```{r}
table(result$cluster, iris$Species)
```
Only 6 measurements have been misclassified.

```{r}
qplot(Petal.Width, Petal.Length, data=iris, color=Species)
```

```{r}
qplot(Petal.Width, Petal.Length, data=iris, color=factor(result$cluster))
```