---
title: "Exploration of Movies Dataset"
author: "Shailesh Kumar"
output: 
  html_notebook:
    toc: true
---

```{r}
require(ggplot2)
require(ggplot2movies)
```
```{r}
head(movies)
```

The movie genres are stored as Boolean flag columns in this dataset.
This is useful since one movie may fall into multiple genres. 

Let's prepare a simpler data set consisting of movie budget, year 
and the short flag 
in which there is a separate column
called genre. Since one movie may fall into multiple genres,
our dataset may have repeated entries for same movie.
This is slightly involved.
```{r}
my.columns <- c('budget', 'year', 'Short')
action.movies <- movies[movies$Action == 1, my.columns]
action.movies$type <- 'action'
animation.movies <- movies[movies$Animation == 1, my.columns]
animation.movies$type <- 'animation'
comedy.movies <- movies[movies$Comedy == 1, my.columns]
comedy.movies$type <- 'comedy'
drama.movies <- movies[movies$Drama == 1, my.columns]
drama.movies$type <- 'drama'
documentary.movies <- movies[movies$Documentary == 1, my.columns]
documentary.movies$type <- 'documentary'
romance.movies <- movies[movies$Romance == 1, my.columns]
romance.movies$type <- 'romance'
movie.budgets <- rbind(action.movies, animation.movies, comedy.movies,
      drama.movies, documentary.movies, romance.movies)
names(movie.budgets) <- c('budget', 'year', 'short', 'type')
head(movie.budgets)
```

Budget information is missing for most movies
```{r}
table(is.na(movie.budgets$budget))
```

Let's see the distribution of movie genres:
```{r}
qplot(type, data=movie.budgets, geom='bar', fill=type)
```

Let's see the breakup of movies by their genre between short and long ones:
```{r}
qplot(type, data=movie.budgets, geom='bar', fill=factor(short))
```

Animation movies are usually short it appears. 

