---
title: "Exploration of MPG Dataset"
author: "Shailesh Kumar"
output: 
  html_notebook:
    toc: true
---

This notebook explores the mpg dataset available in the ggplot2 package. 
The goals of this notebook are:

* General understanding of mpg dataset
* Learning features of ggplot2 package

```{r}
require(ggplot2)
```
This dataset provides fuel economy data from 1999 and 2008 for 
38 popular models of cars. The dataset is shipped with *ggplot2* package.

| Variable | Type | Description | Details | 
|----------|------|-------------|---------|
| manufacturer | string | car manufacturer | 15 manufacturers |
| model | string |model name | 38 models |
| displ | numeric | engine displacement in liters| 1.6 - 7.0, median: 3.3 |
| year | integer | year of manufacturing | 1999, 2008 |
| cyl |  | number of cylinders | 4, 5, 6, 8 | 
| trans| string | type of transmission | automatic, manual (many sub types) |
| drv | string | drive type | f, r, 4, f=front wheel, r=rear wheel, 4=4 wheel|
| cty | integer | city mileage | miles per gallon |
| hwy | integer | highway mileage | miles per gallon |
| fl | string | fuel type | 5 fuel types (diesel, petrol, electric, etc.) |
| class | string | vehicle class | 7 types (compact, SUV, minivan etc.) | 
| |||

Description of mpg dataset
```{r}
str(mpg)
```

Number of rows or observations or records
```{r}
nrow(mpg)
```

Number of columns or variables
```{r}
ncol(mpg)
```

Names of columns
```{r}
colnames(mpg)
```

### Manufacturers
```{r}
table(mpg$manufacturer)
qplot(manufacturer, data=mpg, geom="bar", fill=manufacturer)
```

### Year of manufacturing
```{r}
table(mpg$year)
qplot(factor(year), data=mpg, geom="bar", fill=factor(year))
```


### Displacement
```{r}
summary(mpg$displ)
qplot(displ, data=mpg, geom="histogram", bins=30)
boxplot(mpg$displ)
```

### Cylinders
```{r}
table(mpg$cyl)
qplot(cyl, data=mpg, geom="bar", fill=factor(cyl))
```

### Transmission types
```{r}
table(mpg$trans)
qplot(trans, data=mpg, geom="bar", fill=factor(trans))
```

### Wheel drive types
```{r}
table(mpg$drv)
qplot(drv, data=mpg, geom="bar", fill=drv)
```

### Mileage in city

```{r}
summary(mpg$cty)
qplot(cty, data=mpg, geom="histogram", bins=20)
qplot(cty, data=mpg, geom="histogram", bins=30)
qplot(cty, data=mpg, geom="histogram", bins=40)
```

### Mileage in highways
```{r}
summary(mpg$hwy)
qplot(hwy, data=mpg, geom="histogram", bins=20)
qplot(hwy, data=mpg, geom="histogram", bins=30)
qplot(hwy, data=mpg, geom="histogram", bins=40)
```

### Fuel types
```{r}
table(mpg$fl)
qplot(fl, data=mpg, geom="bar", fill=fl)
```

### Vehicle Class
```{r}
table(mpg$class)
qplot(class, data=mpg, geom="bar", fill=class)
```

### Displacement vs highway efficiency
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color='blue')
```

The same graph can be built using qplot too
```{r}
qplot(displ, hwy, data=mpg, geom="point", color='red')
```

Looking at this data separately for each class
```{r}
qplot(displ, hwy, data=mpg, geom="point", color=class)
```

Fixing the color for all points
```{r}
qplot(displ, hwy, data=mpg, geom="point", color=I("blue"))
```

The same graphic through ggplot
```{r}
ggplot(data=mpg) + geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```


Separate graphs for each vehicle class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=class)) +
  facet_wrap(~ class, nrow = 2)
```

Creating facets on the basis of two variables : number of cylinders and
type of drive
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=drv)) +
  facet_grid(drv ~ cyl)
```

There are no cars for some combinations of drive and cylinders. 

Creating facets vertically on drive:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=drv)) +
  facet_grid(drv ~ .)
```

Creating facets horizontally on cylinders:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=cyl)) +
  facet_grid(. ~ cyl)
```


Estimating a smooth curve for the relationship between displacement and
highway mileage:

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Separate curve for each type of drive:
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, color=drv))
```

Grouping data:
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

Grouping data using specific color:

```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv)
  )
```

Hiding the legend:
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE
  )
```

Overlaying a smooth curve on top of scatter plot:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth()
```

Using a color based grouping for the scatter plot but a common
curve overlayed on top of it:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

Filtering data for a specific geom:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = dplyr::filter(mpg, class == "subcompact"), se = FALSE)
```
The se=FALSE setting removes the confidence interval around the
estimated curve.

Grouping data by drive and then drawing scatter plot with estimated curve
for each group:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```

### Drive vs Cylinder

Visualizing the combinations of drives and cylinders available in the dataset:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(y = drv, x = cyl, color=factor(cyl)), size=4)
```

### Is automatic transmission better? 

We will create a copy of our data frame
```{r}
mpg2 <- mpg
```

We will introduce a new column defining whether the car
is automatic or manual
```{r}
mpg2$is.automatic <- startsWith(mpg2$trans, 'auto')
mpg2$transmission <- ifelse(mpg2$is.automatic, 'auto', 'man')
```
Let's verify the statistics
```{r}
table(mpg2$trans)
table(mpg2$is.automatic)
```


Let's compare the box plots of city mileage for the two types
```{r}
qplot(transmission, cty, data=mpg2, geom='boxplot', fill=transmission, color=transmission)
```
This graphic suggests that the manual transmission is better
than automatic one.

It is time for us to perform a t-test to verify the accuracy.
```{r}
manual.cty <- mpg2$cty[!mpg2$is.automatic]
auto.cty <- mpg2$cty[mpg2$is.automatic]
t.test(manual.cty, auto.cty, alternative = "two.sided", var.equal = FALSE)
```

The p-value is very low.  It indicates strong evidence against the null hypothesis, so we reject the null hypothesis and accept that
means are not equal.

Let's also verify the hypothesis that mean of manual is
greater than mean of automatic transmission.

```{r}
t.test(manual.cty, auto.cty, alternative = "greater", var.equal = FALSE)
```
Again we see that p-value is very very small. Thus, the
alternate hypothesis must be true. Mean of manual
transmission is greater than mean of automatic transmission.

```{r}
t.test(manual.cty, auto.cty, alternative = "less", var.equal = FALSE)
```

Here p-value is 1. We don't have any evidence supporting the
alternate hypothesis.

