---
title: "Exploration of Wine Dataset"
author: "Shailesh Kumar"
output: 
  html_notebook:
    toc: true
    number_sections: true
---
# Overview

The *wine* dataset is available in *rattle.data* package. it contains
the results of a chemical analysis of wine grown in specific area
of Italy. There are 178 records and 14 variables. 

| Variable | Type | Description |
|----------|------|-------------|
| Type | factor | Wine type: 1, 2, 3 |
| Alcohol | | |
| Malic | | Malic acid |
| Ash |  | Ash |
| Alcalinity | | Alcalinity of ash |
| Magnesium |  | Magnesium |
| Phenols | | Total phenols |
| Flavanoids | | Flavanoids |
| Nonflavanoids | | Non-flavanoid phenols |
| Proanthocyanins | | |
| Color | | Color intensity | 
| Hue | | Hue |
| Dilution |  | D280/OD315 of diluted wines |
| Proline | | Proline |



Loading required libraries
```{r}
require(rattle.data)
require(ggplot2)
require(corrplot)
require(cluster)
```


Loading the data into search path
```{r}
data(wine, package='rattle.data')
```


```{r}
head(wine)
```

```{r}
dim(wine)
```

# Variables

## Type

```{r}
summary(wine$Type)
```

## Alcohol

```{r}
summary(wine$Alcohol)
```

```{r}
qplot(Type, Alcohol, data=wine, geom='boxplot', color=Type)
```

## Malic

```{r}
summary(wine$Malic)
```


```{r}
qplot(Type, Malic, data=wine, geom='boxplot', color=Type)
```

## Ash

```{r}
summary(wine$Ash)
```

```{r}
qplot(Type, Ash, data=wine, geom='boxplot', color=Type)
```

## Alcalinity

```{r}
summary(wine$Alcalinity)
```

```{r}
qplot(Type, Alcalinity, data=wine, geom='boxplot', color=Type)
```

## Magnesium

```{r}
summary(wine$Magnesium)
```

```{r}
qplot(Type, Magnesium, data=wine, geom='boxplot', color=Type)
```

## Phenols

```{r}
summary(wine$Phenols)
```

```{r}
qplot(Type, Phenols, data=wine, geom='boxplot', color=Type)
```

## flavanoids

```{r}
summary(wine$Flavanoids)
```

```{r}
qplot(Type, Flavanoids, data=wine, geom='boxplot', color=Type)
```

## Non Flavanoids
```{r}
summary(wine$Nonflavanoids)
```

```{r}
qplot(Type, Nonflavanoids, data=wine, geom='boxplot', color=Type)
```

## Proanthocyanins

```{r}
summary(wine$Proanthocyanins)
```

```{r}
qplot(Type, Proanthocyanins, data=wine, geom='boxplot', color=Type)
```

## Color

```{r}
summary(wine$Color)
```

```{r}
qplot(Type, Color, data=wine, geom='boxplot', color=Type)
```

## Hue

```{r}
summary(wine$Hue)
```

```{r}
qplot(Type, Hue, data=wine, geom='boxplot', color=Type)
```

## Dilution

```{r}
summary(wine$Dilution)
```

```{r}
qplot(Type, Dilution, data=wine, geom='boxplot', color=Type)
```

## Proline

```{r}
summary(wine$Proline)
```

```{r}
qplot(Type, Proline, data=wine, geom='boxplot', color=Type)
```

# Correlations

```{r}
wine.cor <- cor(wine[, -1])
wine.cor[wine.cor == 1] <- 0
corrplot(wine.cor, method='circle')
wine.strong <- abs(wine.cor)
wine.strong[wine.strong < 0.6] <- 0
corrplot(wine.strong)
```


# Clustering

Our goal is to predict the type of wine from the remaining
data.

Let us scale the remaining variables:

```{r}
wine.scaled <- scale(wine[-1])
```

## K-means clustering


We will cluster the data into 3 groups.

```{r}
kmeans.result <- kmeans(wine.scaled, centers=3)
```

The centroids:

```{r}
kmeans.result$centers
```

The cluster labels
```{r}
kmeans.result$cluster
```

Size of each cluster
```{r}
kmeans.result$size
```
This can be computed as follows too:
```{r}
table(kmeans.result$cluster)
```

Within cluster sum of squares
```{r}
kmeans.result$withinss
```

Total within cluster sum of squares
```{r}
kmeans.result$tot.withinss
sum(kmeans.result$withinss)
```

Between cluster sum of squares
```{r}
kmeans.result$betweenss
```

Total sum of squares
```{r}
kmeans.result$totss
```


Let us map it to actual type labels to compute the confusion matrix:
```{r}
table(kmeans.result$cluster, wine$Type)
```

It seems that there are 6 samples which have been
incorrectly clustered. Otherwise, the clustering has
delivered very good results.

We can visualize the results of clustering in 2 dimensions. 
This is achieved as follows:

* The data-set is projected into two dimensions using 
  principal component analysis.
* The projected points are visualized on the graph with
  ellipses surrounding each cluster.
* The ellipse boundaries can be colored.
```{r}
clusplot(wine.scaled, kmeans.result$cluster, color=T, shade=T)
```

## Hierarchical clustering

This works on the distances between points.

Let us compute the Euclidean distances:

```{r}
distances.l2 <- dist(wine.scaled, method='euclidean')
```

Let us now run hierarchical clustering
using Ward's minimum variance criterion:
```{r}
hclust.result <- hclust(distances.l2, method='ward.D')
```

The clustering result can be visualized in the form of
a dendogram:
```{r}
plot(hclust.result)
```

The tree can be cut into clusters as follows:
```{r}
hclust.cluster <- cutree(hclust.result, k=3)
```
Let us look at the cluster labels:
```{r}
table(hclust.cluster)
```

Let us prepare our confusion matrix:
```{r}
table(hclust.cluster, wine$Type)
```
A total of 7 points seem to have been misclustered. 

