---
title: "Activation Functions"
author: "Shailesh Kumar"
output: 
  html_notebook:
    toc: true
    number_sections: true
---

This notebook contains descriptions of several activation functions.


Necessary packages
```{r}
require(ggplot2)
```

# Identity Function

```{r}
f <- function(x) {x}
ggplot(data.frame(x=c(-4, 4)), mapping=aes(x=x)) + 
  geom_hline(yintercept=0, color='red') +
  geom_vline(xintercept=0, color='red') +
  stat_function(fun=f, colour = "dodgerblue3") +
  ggtitle('Identity') +
  scale_x_continuous(name='x') +
  scale_y_continuous(name='f(x)') +
  theme(plot.title = element_text(hjust = 0.5))

```

Let's create a reusable function for plotting the activation functions going
forward:
```{r}
plot_activation_function <- function(f, title, range){
  start <- range[1]
  end <- range[2]
  ggplot(data.frame(x=range), mapping=aes(x=x)) + 
    geom_hline(yintercept=0, color='red', alpha=1/4) +
    geom_vline(xintercept=0, color='red', alpha=1/4) +
    stat_function(fun=f, colour = "dodgerblue3") +
    ggtitle(title) +
    scale_x_continuous(name='x') +
    scale_y_continuous(name='f(x)') +
    theme(plot.title = element_text(hjust = 0.5))
}
```

# Binary Step Function
```{r}
f <- function(x){
  ifelse(x >= 0, 1, 0)
}
plot_activation_function(f, 'Binary step', c(-4,4))
```


# Logistic Function
```{r}
f <- function(x){1 / (1 + exp(-x))}
plot_activation_function(f, 'Logistic', c(-4,4))
```

# Tan H Function

```{r}
tanh_func <- function(x){tanh(x)}
plot_activation_function(tanh_func, 'TanH', c(-4,4))
```

# Arc Tan Function

```{r}
arc_tan_func <- function(x){atan(x)}
plot_activation_function(arc_tan_func, 'ArcTan', c(-4,4))
```

# Soft Sign Function
```{r}
soft_sign_func <- function(x){ x / (1 + abs(x))  }
plot_activation_function(soft_sign_func, 'Softsign', c(-4,4))
```


# Rectified Linear Unit Function
```{r}
rec_lu_func <- function(x){ ifelse(x < 0 , 0, x )}
plot_activation_function(rec_lu_func, 'RecLU', c(-4,4))
```

# Leaky Rectified Linear Unit Function
```{r}
leaky_rec_lu_func <- function(x){ ifelse(x < 0 , 0.01 *x , x )}
plot_activation_function(leaky_rec_lu_func, 'Leaky RecLU', c(-8,8))
```

# Soft Plus Function
```{r}
soft_plus_func <- function(x){ log(1 + exp(x))}
plot_activation_function(soft_plus_func, 'SoftPlus', c(-8,8))
```

# Bent Identity Function
```{r}
bent_identity_func <- function(x){  (sqrt(x^2 + 1) - 1)/2 + x  }
plot_activation_function(bent_identity_func, 'Bent Identity', c(-8,8))
```

# Sinusoid Function
```{r}
plot_activation_function(sin, 'sin(x)', c(-8,8))
```

# Sinc Function
```{r}
sinc_function<- function(x){ifelse(x == 0, 1, sin(x) / x)}
plot_activation_function(sinc_function, 'Sinc', c(-8,8))
```


# Gaussian Function
```{r}
gaussian_function<- function(x){exp(-x^2)}
plot_activation_function(gaussian_function, 'Gaussian', c(-4,4))
```















